---
title: "Assignment 1 FoML"
author: "Zach Ours"
date: "2026-09-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup
For this assignment, I will use NFL play-by-play data from the nflverse project.I chose this dataset because I am interested in sports analytics and wanted to explore the relationship between different statistics from NFL plays.

```{r}
library(nflreadr)
library(dplyr)
```



```{r}
# Source: nflverse NFL play-by-play data
nfl_data <- load_pbp(2025)
```


## View the Data

```{r}
head(nfl_data)
```


```{r}
dim(nfl_data)
```


## Select Variables

```{r}
nfl_small <- nfl_data %>%
  select(
    posteam,
    defteam,
    play_type,
    yards_gained,
    air_yards,
    yards_after_catch,
    down
  )
```


```{r}
head(nfl_small)
```



## Descriptive Statistics

### Quantitative Variables

```{r}
summary(nfl_small$yards_gained)
```

```{r}
summary(nfl_small$air_yards)
```


### Categorical Variables

```{r}
table(nfl_small$play_type)
```

```{r}
table(nfl_small$posteam)
```


## Variable Transformation

A new variable called `big_play` was created. A play is classified as a big play if it gained at least 20 yards.

```{r}
nfl_small <- nfl_small %>%
  mutate(
    big_play = ifelse(yards_gained >= 20, "Yes", "No")
  )
```


```{r}
table(nfl_small$big_play)
```


## Histogram

The following histogram shows the distribution of yards gained per play. Most plays fall within a relatively small range of yards gained, while larger gains occur less frequently.

```{r}
hist(
  nfl_small$yards_gained[nfl_small$yards_gained >= -20 &
                           nfl_small$yards_gained <= 50],
  main = "Distribution of Yards Gained",
  xlab = "Yards Gained"
)
```


## Scatterplot

The following scatterplot compares air yards to total yards gained.

```{r}
plot(
  nfl_small$air_yards,
  nfl_small$yards_gained,
  main = "Air Yards vs. Yards Gained",
  xlab = "Air Yards",
  ylab = "Yards Gained",
  pch = 19
)
```

The scatterplot shows a general positive relationship between air yards and total yards gained. Passes thrown farther downfield often result in more total yards, although there is considerable variation between individual plays.

## Conclusion

In this assignment, I used R to analyze NFL play-by-play data from the 2025 season. I selected quantitative and categorical variables, calculated descriptive statistics, transformed the yards gained variable to identify big plays, and created a histogram and scatterplot to visualize the data.